---
title: "ETC5521 Assignment 1"
subtitle: "House and Mortgage Data"
team: Dugong
author:
  - Abhishek Sinha
  - Yezi He
  - Yawen Zhang
  - Cuiping Wei
output: 
  bookdown::html_document2
bibliography: reference.bib
biblio-style: authoryear-comp
link-citations: yes
linestretch: 3
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2: 
    toc: yes
    toc_float:
      collapsed: true
      number_sections: false
    theme: paper
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE,message = FALSE)
library(tidyverse)
library(scales)
library(lubridate)
library(broom)
library(MASS)
library(stringr)
library(kableExtra)
library(readxl)
library(janitor)
library(rvest)
library(dplyr)
library(plotly)
```

```{r write-data}
# state_hpi <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-05/state_hpi.csv")
# mortgage_rates <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-05/mortgage.csv")
# recession_dates <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-05/recessions.csv")

# write_csv(state_hpi, here::here("data", "state_hpi.csv"))
# write_csv(mortgage_rates, here::here("data", "mortgage_rates.csv"))
# write_csv(recession_dates, here::here("data", "recession_dates.csv"))
```

```{r read-data}
state_hpi <-readr::read_csv(here::here("data/state_hpi.csv")) 
mortgage_rates<-readr::read_csv(here::here("data/mortgage_rates.csv")) 
recession_dates<-readr::read_csv(here::here("data/recession_dates.csv")) 
```


# Introduction and motivation


# Data description




# Analysis and findings


## Can HPI and mortgage rates provide evidence of a subprime crisis? 

```{r recession-wrangle}
# separate the period time to "start" and "end"
recession_dat <- recession_clean  %>% 
  separate(period_range, into = c("start", "end"), sep = "–") %>% 
  mutate(start = as.Date(parse_date_time(start, "%b %Y")), 
         end = as.Date(parse_date_time(end, "%b %Y"))) 

recession_tab <- recession_dat %>% 
  select(1:4,6) %>% 
  kable(caption = "Early recessions and crises") %>% 
  kable_styling()
recession_tab
```


```{r}
state_hpi <- state_hpi %>% 
  mutate(date = ymd(year*10000+month*100+1)) 

nation_hpi <- state_hpi %>%
  #mutate(date = ymd(year*10000+month*100+1)) %>% 
  dplyr::select("date","year","month","us_avg") %>%
  distinct()

```


```{r miss, fig.cap="Missing value for the mortgage rates, only available for fixed 30 years don't have NA"}
mortgage_rates %>%
  select(2,4,6) %>% 
  vis_miss()
```


```{r state_abb_data}
# data of 2-letter abbreviations for the state names
state_abb <- as.data.frame(state.abb)
state_abb <- state_abb %>% mutate(region = state.region)
colnames(state_abb)[1] <- "state"

# join the state HPI data
state_hpi_abb <- merge(x=state_hpi, y=state_abb, by="state", all.x = TRUE)
state_hpi_abb$region[is.na(state_hpi_abb$region)] <- "South"

# divided HPI data into four regions, choose median represents region HPI
state_data_hpi <- state_hpi_abb %>% 
  dplyr::select("date", "region", "state", "price_index", "us_avg") %>% 
  group_by(date,region,us_avg) %>% 
  summarise(price_index = median(price_index))
```

```{r hpi-mort}
# national HPI
hpi_mort <- ggplot()+ 
  geom_line(data=mortgage_rates, 
            aes(x=date, y=fixed_rate_30_yr*10), 
            color="red") +
  geom_line(data=nation_hpi, 
            aes(x=date, y=us_avg), 
            color="green")+
  scale_y_continuous(sec.axis = sec_axis(~ . *0.1, name = "Mortgage Rates"))+
  xlim(ymd(19750101), NA) +
  labs(x="Date", y="US Average HPI") +
  ggtitle("HPI vs. Mortgage rates")

```

```{r nation_recession, fig.cap="The trend for Recession, National HPI and Mortgage rates. The red line represents the mortgage rates, the green line represents HPI, and the shaded blue area represents the duration of the recession."}
# combine national HPI and recession into one plot
mortgage_recession <- hpi_mort +
  geom_rect(data = recession_dat, 
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
            alpha = 0.25, 
            fill = "blue") +
  ggtitle("Recession, HPI and Mortgage Rates")+
  theme_light()

mortgage_recession

```


## Can we find any house price bubble in the U.S from 1975 to 2018?

```{r bubble, fig.cap="Recession and National HPI. The occurrence of the US housing bubble from 1975 to 2018 and the scenarios that may present in the future."}
# bubble plot
p_bubble <- state_hpi %>% 
  ggplot() +
  geom_line(aes(date, us_avg), 
            color = "green") +
  geom_rect(data = recession_dat, 
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
            alpha = 0.25, fill = "blue") +
  annotate("segment", 
           x=as.Date('2001-02-01'), 
           xend=as.Date('2012-01-01'), 
           y=100, yend=100, 
           colour = "red", 
           size = 0.6)+
  annotate("segment", 
           x=as.Date('2012-01-01'), 
           xend=as.Date('2023-09-01'), 
           y=120, yend=120, 
           colour = "red", 
           size = 0.6)+
  annotate("pointrange", 
           x = as.Date('2006-12-01'), 
           y = 166, 
           ymin = 100, ymax = 190, 
           colour = "red", 
           size = 0.6)+
  annotate("pointrange", 
           x = as.Date('2019-01-01'), 
           y = 190, 
           ymin = 120, ymax = 200, 
           colour = "red", 
           size = 0.6)+
  annotate("segment", 
           x = as.Date('2019-01-01'), 
           xend=as.Date('2025-09-01'), 
           y = 192, yend = 100, 
           colour = "#69b3a2", 
           size=1, alpha=1, 
           arrow=arrow())+
  annotate(geom = "text", 
           x=as.Date('2025-12-01'), 
           y=85, 
           label="Future\nscenario ?", 
           size = 4, 
           color = "red")+
  annotate(geom = "text", 
           x=as.Date('2006-12-01'), 
           y=90, 
           label="Bubble1", 
           size = 4, 
           color = "red")+
  annotate(geom = "text", 
           x=as.Date('2019-01-01'), 
           y=110, 
           label="Bubble2", 
           size = 4, 
           color = "red")+
  ggtitle("Recession, National HPI and Housing Bubble")+
  xlim(c(ymd(19750101),ymd(20261201))) +
  ylab("US Average HPI") +
  xlab("Date") +
  theme_light()

p_bubble
```


## Was the annual change in HPI index similar to the annual change in mortgage rates?  

```{r mort-fixed30}
mortgage_fix30 <- mortgage_rates %>% 
  select(1:2) %>% 
  mutate(year = year(date)) %>% 
  filter(year >= 1975) %>% 
  group_by(year) %>% 
  summarise(fixed_30yr_avg = mean(fixed_rate_30_yr)) %>% 
  mutate(ratio = (fixed_30yr_avg - lag(fixed_30yr_avg)) / fixed_30yr_avg,
         ratio = round(ratio,3))

```

```{r nation-hpi-change}
nation_hpi_change <- nation_hpi %>% 
  group_by(year) %>% 
  summarise(us_avg = mean(us_avg)) %>% 
  mutate(ratio = (us_avg - lag(us_avg)) / us_avg,
         ratio = round(ratio,3))
  
```

```{r change, fig.cap="**Mortgage annual changes** VS **HPI annual changes**"}
p1<-mortgage_fix30 %>% 
  ggplot(aes(year, ratio)) + 
  geom_bar(aes(fill=factor((ratio>0)+1)),
           stat="identity")+
  scale_x_continuous(breaks = seq(1975,2020,6))+
  ggtitle("Mortgage annual changes")+
  # scale_fill_discrete(name="Change",
  #                        breaks=c("1", "2"),
  #                        labels=c("Down", "Up"))+
  scale_fill_manual("change",
                    breaks=c("1","2"), 
                    labels=c("Down", "Up"),
                    values=c("#FFAEB9","#7AC5CD"))+
  guides(fill = guide_legend(reverse=TRUE))+
  theme_light()

p2<-nation_hpi_change %>% 
  ggplot(aes(year, ratio)) + 
  geom_bar(aes(fill=factor((ratio>0)+1)),
           stat="identity") +
  scale_x_continuous(breaks = seq(1975,2020,6))+
  ggtitle("HPI annual changes")+
  # scale_fill_discrete(name="Change",
  #                        breaks=c("1", "2"),
  #                        labels=c("Down", "Up"))+
   scale_fill_manual("change", 
                     breaks=c("1","2"),
                     labels=c("Down", "Up"),
                    values=c("#FFAEB9","#7AC5CD"))+
  guides(fill = guide_legend(reverse=TRUE))+
  theme_light()

p1/p2+plot_layout(guides = "collect")

```


# Conclusion


# Acknowledgments
The following packages are used to produce this report:
tidyverse [@tidyverse], lubridate [@lubridate], stringr [@stringr], kableExtra [@kableExtra], readxl [@readxl], rvest [@rvest], janitor [@janitor], dplyr [@dplyr] , plotly [@plotly], MASS[@mass], scales[@scales], broom[@broom]

# References
